% ============================================================================
%  SECTION 3 — LECTURE NOTES
%  The Philosophy of Machine Minds
%  Comprehensive instructor notes for presentation delivery
% ============================================================================
\documentclass[11pt, a4paper]{article}

\usepackage[margin=2.5cm]{geometry}
\usepackage{amsmath, amssymb, amsthm}
\usepackage{tikz}
\usetikzlibrary{arrows.meta, positioning, calc}
\usepackage{booktabs}
\usepackage{enumitem}
\usepackage{hyperref}
\usepackage[dvipsnames]{xcolor}
\usepackage{tcolorbox}
\usepackage{microtype}
\usepackage{parskip}
\usepackage{titlesec}
\usepackage{fancyhdr}
\usepackage{marginnote}

% --- Colours ---
\definecolor{accent}{HTML}{BF360C}
\definecolor{muted}{HTML}{546E7A}
\definecolor{boxbg}{HTML}{FFF3E0}
\definecolor{defbg}{HTML}{E3F2FD}
\definecolor{warnbg}{HTML}{FCE4EC}

% --- Boxes ---
\tcbuselibrary{breakable, skins}

\newtcolorbox{keyinsight}[1][]{%
  colback=boxbg, colframe=accent, fonttitle=\bfseries,
  title={Key Insight}, breakable, #1}

\newtcolorbox{definition}[1][]{%
  colback=defbg, colframe=MidnightBlue, fonttitle=\bfseries,
  title={Definition}, breakable, #1}

\newtcolorbox{speakingnote}[1][]{%
  colback=gray!8, colframe=muted, fonttitle=\bfseries\color{muted},
  title={Speaking Note}, breakable, left=4mm, right=4mm, #1}

\newtcolorbox{warning}[1][]{%
  colback=warnbg, colframe=Red, fonttitle=\bfseries,
  title={Caution}, breakable, #1}

\newtcolorbox{example}[1][]{%
  colback=green!5, colframe=OliveGreen, fonttitle=\bfseries,
  title={Example}, breakable, #1}

% --- Headers ---
\pagestyle{fancy}
\fancyhf{}
\fancyhead[L]{\small\textsc{Section 3 --- The Philosophy of Machine Minds}}
\fancyhead[R]{\small\thepage}
\renewcommand{\headrulewidth}{0.4pt}

% --- Title formatting ---
\titleformat{\section}{\Large\bfseries\color{accent}}{}{0em}{}[\vspace{-0.3em}\rule{\textwidth}{0.8pt}]
\titleformat{\subsection}{\large\bfseries}{}{0em}{}

% --- Meta ---
\title{\textbf{Lecture Notes}\\[0.3em]
  \LARGE The Philosophy of Machine Minds\\[0.2em]
  \large Prediction, Representation, and Understanding\\
  in the Age of Large Language Models}
\author{Section 3 --- Executive AI Programme}
\date{2026}

% ============================================================================
\begin{document}
\maketitle
\thispagestyle{empty}

\vspace{1em}
\begin{center}
\begin{tcolorbox}[colback=gray!5, colframe=black, width=0.85\textwidth]
\textbf{Session duration:} 30 minutes (25 min presentation + 5 min discussion)\\
\textbf{Narrative arc:} Mathematical identity $\to$ cosmic empirical observation $\to$ oldest open question in philosophy of mind\\
\textbf{Unifying thread:} \emph{LLMs are forcing us to redefine what it means to understand.}
\end{tcolorbox}
\end{center}

\tableofcontents
\newpage

% ============================================================================
\section{Opening: The Arc of the Argument}
% ============================================================================

This session is structured as a single argument in three acts. Each act
builds on the previous one, and the conclusion of the third act depends on
having internalised the first two.

\textbf{Act I} (Topic A) establishes a mathematical identity: prediction,
compression, and intelligence are the same operation, viewed from different
angles. This is not metaphor --- it is a theorem-level claim traceable to
Shannon (1948), Kolmogorov (1963), Solomonoff (1964), and Hutter (2005).

\textbf{Act II} (Topic B) presents an empirical observation that follows
naturally from Act I: if all powerful learning systems are compressing
reality, they should converge on similar internal representations. The
Platonic Representation Hypothesis (Huh et al., ICML 2024) provides
evidence that this is exactly what is happening.

\textbf{Act III} (Topic C) confronts the philosophical consequence: if
models compress, build world models, and converge on reality's structure,
can we still deny that they ``understand''? The Chinese Room argument
(Searle, 1980) says yes. Recent empirical evidence says: it's complicated.

\begin{speakingnote}
Open with energy and promise. The audience should feel that this session
will change how they think about a fundamental concept. Say something like:
``This session is different from what came before. We've seen what LLMs can
do. Now we confront what those capabilities \emph{mean}. By the end, you
may find that the word `understanding' --- a word you use fifty times a
day --- no longer means what you thought it did.''
\end{speakingnote}

% ============================================================================
\section{Topic A: Prediction = Compression = Intelligence}
% ============================================================================

\subsection{Opening Hook: Compression as Understanding}

The central claim of Topic A is that \textbf{compression and understanding
are the same thing}. Before making this rigorous, we build intuition through
a series of examples. In each one, notice that ``understanding'' a pattern
is equivalent to finding a short rule that reproduces it.

\begin{example}[title={Example 1: The Fibonacci Sequence}]
Display the sequence: $1, 1, 2, 3, 5, 8, 13, \;?$

You replaced an infinite list with a two-word rule: \emph{``add the
previous two.''} Formally, the rule is $a_n = a_{n-1} + a_{n-2}$, with
$a_1 = a_2 = 1$. This rule is far shorter than the sequence it generates.
That act of compression \textbf{is} understanding.
\end{example}

\begin{example}[title={Example 2: A Coin Toss vs.\ a Heartbeat}]
Consider two sequences of 1,000 numbers:
\begin{itemize}[leftmargin=2em]
  \item \textbf{Sequence A:} Fair coin tosses: $0, 1, 1, 0, 1, 0, 0, 1, 0, 1, \ldots$ --- There is no pattern. Each bit is independent. You cannot do better than recording all 1,000 bits. This sequence is \emph{incompressible}.
  \item \textbf{Sequence B:} A resting heartbeat: $72, 71, 73, 72, 71, 72, 73, \ldots$ --- The values hover around 72 with small fluctuations. You could compress this to: ``$72 \pm$ small noise.'' Perhaps 50 bits suffice instead of 1,000. The sequence is highly \emph{compressible}.
\end{itemize}
The compressible sequence is the one you ``understand'' --- you can describe its generating process. The incompressible one is the one that surprises you at every step.
\end{example}

\begin{example}[title={Example 3: Language Itself}]
Consider the partial sentence: ``The cat sat on the \underline{\hspace{1.5cm}}.'' You can predict with high confidence that the next word is ``mat'' (or ``floor,'' ``chair,'' etc.). Your ability to predict reflects your knowledge of English grammar, common collocations, and the world. If you can predict the next word well, you need fewer bits to encode it --- you have compressed the language.

Now consider: ``Colourless green ideas sleep \underline{\hspace{1.5cm}}.'' If you know Chomsky's famous example, you predict ``furiously.'' If not, the word is surprising --- you need more bits. Your background knowledge (or lack thereof) directly determines your compression rate.
\end{example}

\begin{example}[title={Example 4: Stock Prices}]
A sequence of daily closing prices for a stock: \$142, \$145, \$143, \$148, \ldots If a model can predict tomorrow's price better than chance, it can compress the price sequence. To the extent that financial analysts ``understand'' a company's trajectory, they are compressing its price movements into a narrative (``growth stock with seasonal dips''). An analyst who understands the company better compresses its price history into a shorter description --- and makes better predictions.
\end{example}

\begin{keyinsight}
In every example, the pattern is the same: \textbf{to understand a
sequence is to find a short rule that generates it; to find a short rule
is to compress it; to compress it is to predict the next element}. These
three operations --- understanding, compression, prediction --- are not
merely analogous. As we will now show, they are mathematically identical.
\end{keyinsight}

\begin{speakingnote}
Do not rush past these examples. They are doing real intellectual work ---
establishing the intuition that compression and understanding are not
merely related, but identical. Everything that follows is a formalisation
of what the audience just experienced in their own heads. Let the audience
try to compress each example themselves before revealing the rule.
\end{speakingnote}


\subsection{Shannon's Foundational Result (1948)}

We now make the prediction--compression equivalence precise. This requires
a careful setup of notation and definitions.

\subsubsection{Notation and Setup}

We consider a sequence of symbols $x_1, x_2, x_3, \ldots, x_T$ drawn from
a finite \textbf{alphabet} $\mathcal{A}$. For natural language, $\mathcal{A}$
might be the set of all words in English (or, at a lower level, the set of
all byte values $\{0, 1, \ldots, 255\}$). For a coin toss, $\mathcal{A} =
\{0, 1\}$.

We use the following notational conventions throughout:
\begin{itemize}[leftmargin=2em]
  \item $x_t$ denotes the $t$-th symbol in the sequence, where
    $t \in \{1, 2, \ldots, T\}$.
  \item $x_{<t}$ denotes the \textbf{prefix} (history) of the sequence up
    to but not including position $t$: that is, $x_{<t} = (x_1, x_2,
    \ldots, x_{t-1})$. For $t = 1$, we define $x_{<1} = \emptyset$ (the
    empty sequence).
  \item $x_{\leq t}$ denotes the prefix up to and including position $t$:
    $x_{\leq t} = (x_1, x_2, \ldots, x_t)$.
  \item $x_{1:T}$ denotes the full sequence $(x_1, x_2, \ldots, x_T)$.
  \item $p(x_t \mid x_{<t})$ denotes the \textbf{conditional probability}
    of symbol $x_t$ given the entire preceding history $x_{<t}$. This is
    the probability that a particular symbol appears at position $t$,
    given that we have already observed $x_1, \ldots, x_{t-1}$.
  \item $\log_2$ denotes the base-2 logarithm. All information-theoretic
    quantities are measured in \textbf{bits}.
\end{itemize}

\subsubsection{Shannon Entropy}

\begin{definition}[title={Shannon Entropy}]
Let $X$ be a discrete random variable taking values in a finite set
$\mathcal{A} = \{a_1, a_2, \ldots, a_K\}$ with probability mass function
$p$, so that $p(a_k) = \Pr[X = a_k]$ for each $k \in \{1, \ldots, K\}$.
The \textbf{Shannon entropy} of $X$ is defined as:
\[
H(X) \;=\; -\sum_{k=1}^{K} p(a_k) \,\log_2 p(a_k)
\]
with the convention that $0 \cdot \log_2 0 = 0$ (justified by the
limit $\lim_{p \to 0^+} p \log_2 p = 0$).
\end{definition}

\textbf{Interpretation.} $H(X)$ has three equivalent interpretations:
\begin{enumerate}[leftmargin=2em]
  \item \textbf{Uncertainty:} $H(X)$ measures how uncertain we are about
    the outcome of $X$ before observing it. If $X$ is deterministic
    (one outcome has probability 1), then $H(X) = 0$: no uncertainty.
    If $X$ is uniformly distributed over $K$ outcomes, then
    $H(X) = \log_2 K$: maximum uncertainty.

  \item \textbf{Expected surprise:} The quantity $-\log_2 p(x)$ is called
    the \textbf{surprisal} or \textbf{information content} of outcome $x$.
    A very likely event ($p(x) \approx 1$) has low surprisal
    ($-\log_2 p(x) \approx 0$); a rare event ($p(x) \approx 0$) has high
    surprisal. The entropy $H(X)$ is the \emph{expected} surprisal:
    $H(X) = \mathbb{E}[-\log_2 p(X)]$.

  \item \textbf{Optimal compression rate:} $H(X)$ is the minimum average
    number of bits required to encode one draw from $X$ using any lossless
    code. This is Shannon's source coding theorem (stated precisely in the
    appendix). You cannot, on average, use fewer than $H(X)$ bits per
    symbol, and there exist codes (e.g., Huffman codes, arithmetic codes)
    that come arbitrarily close to this limit.
\end{enumerate}

\subsubsection{Conditional Entropy and the Prediction--Compression Link}

For sequences, the relevant quantity is the \textbf{conditional entropy}:

\begin{definition}[title={Conditional Entropy}]
The \textbf{conditional entropy} of $X_t$ given the history $X_{<t}$ is:
\[
H(X_t \mid X_{<t}) \;=\; \mathbb{E}_{x_{<t}}\!\left[\,
  -\sum_{a \in \mathcal{A}} p(a \mid x_{<t}) \,\log_2 p(a \mid x_{<t})
\,\right]
\]
where the outer expectation is over all possible histories $x_{<t}$,
drawn according to their probability under the true data distribution.
\end{definition}

$H(X_t \mid X_{<t})$ measures how many bits you need to encode the $t$-th
symbol, on average, \emph{after} you have already observed the preceding
symbols. The better you can predict $x_t$ from $x_{<t}$, the lower
$H(X_t \mid X_{<t})$ is.

\textbf{The fundamental link.} Shannon's result can now be stated
concisely: for any model $p_\theta$ that assigns conditional probabilities
$p_\theta(x_t \mid x_{<t})$, the optimal code length for symbol $x_t$
given history $x_{<t}$ is:
\[
\ell^*(x_t \mid x_{<t}) \;=\; -\log_2 p_\theta(x_t \mid x_{<t})
\]
This is exact for arithmetic coding (see Section~\ref{sec:arithmetic})
and approximate for other coding schemes. The equation means:
\begin{itemize}[leftmargin=2em]
  \item If $p_\theta(x_t \mid x_{<t})$ is close to 1 (the model is very
    confident in the correct prediction), then $-\log_2 p_\theta \approx 0$
    bits: the symbol costs almost nothing to encode. The model has
    effectively ``known'' what was coming --- it has compressed the data.

  \item If $p_\theta(x_t \mid x_{<t})$ is close to $1/|\mathcal{A}|$ (the
    model is essentially guessing uniformly), then
    $-\log_2 p_\theta \approx \log_2 |\mathcal{A}|$ bits: the symbol
    costs as many bits as if we had no model at all. No compression.

  \item Therefore: \textbf{better prediction $\Leftrightarrow$ shorter
    code $\Leftrightarrow$ better compression.} These are not two related
    tasks --- they are the \emph{same} mathematical operation expressed
    in dual form.
\end{itemize}

\begin{keyinsight}
Any system that does next-token prediction is \emph{implicitly} performing
data compression. When we train an LLM to minimise the cross-entropy
loss, we are literally training it to be a better compressor. The
standard LLM training objective is:
\[
\mathcal{L}(\theta) \;=\; -\frac{1}{T}\sum_{t=1}^{T}
\log_2 p_\theta(x_t \mid x_{<t})
\]
where $\theta$ denotes the model parameters, $T$ is the sequence length,
$x_t$ is the $t$-th token, and $x_{<t}$ is the history. This loss is
\emph{exactly} the average number of bits per token that the model would
use to compress the training data via arithmetic coding. Minimising
$\mathcal{L}(\theta)$ is equivalent to making the model a better
compressor.
\end{keyinsight}

\subsection{The Theoretical Chain: Solomonoff, Kolmogorov, Hutter}

Shannon established that prediction and compression are two views of the
same operation. The next link in the chain connects compression to
intelligence itself. This requires three theorists and three decades.

\subsubsection{Kolmogorov Complexity (1963)}

\begin{definition}[title={Kolmogorov Complexity}]
Fix a universal Turing machine $U$ (a Turing machine that can simulate
any other Turing machine given the appropriate program as input). The
\textbf{Kolmogorov complexity} of a finite binary string $x$, denoted
$K_U(x)$, is the length (in bits) of the shortest program $\pi$ that,
when executed on $U$, outputs exactly $x$ and then halts:
\[
K_U(x) \;=\; \min\bigl\{\,|\pi| \;:\; \pi \in \{0,1\}^*,\;
U(\pi) = x\,\bigr\}
\]
where $|\pi|$ denotes the length of program $\pi$ in bits, $\{0,1\}^*$
is the set of all finite binary strings, and $U(\pi) = x$ means that
$U$ run on input $\pi$ produces output $x$ and halts.
\end{definition}

\textbf{Key properties:}
\begin{enumerate}[leftmargin=2em]
  \item \textbf{Invariance theorem.} For any two universal Turing machines
    $U_1, U_2$, $|K_{U_1}(x) - K_{U_2}(x)| \leq c$ for some constant $c$
    that depends on $U_1, U_2$ but not on $x$. So $K(x)$ is well-defined
    up to an additive constant, and we write $K(x)$ without specifying $U$.

  \item \textbf{Uncomputability.} $K(x)$ is not computable: there exists
    no algorithm that takes $x$ as input and returns $K(x)$. (See proof
    sketch in the appendix.) This is why we use \emph{approximations} to
    $K(x)$ --- gzip, PNG, FLAC, and language models are all compressors
    that approximate $K(x)$ from above.

  \item \textbf{Incompressible strings.} A string $x$ of length $n$ is
    \emph{incompressible} if $K(x) \geq n$: no program shorter than the
    string itself can produce it. A counting argument shows that the
    majority of $n$-bit strings are incompressible (or nearly so). Truly
    random data cannot be compressed.
\end{enumerate}

\subsubsection{Solomonoff Induction (1964)}

Ray Solomonoff asked: what is the \emph{optimal} way to predict the next
symbol in a sequence, if you have unlimited computational resources?
His answer constitutes one of the deepest results in the theory of
inference.

\begin{definition}[title={Solomonoff's Universal Prior}]
Fix a prefix-free universal Turing machine $U$. (A machine is
\emph{prefix-free} if no valid program is a prefix of another valid
program; this is a technical condition ensuring that probabilities sum
to at most 1.) The \textbf{Solomonoff prior} (also called the
\emph{universal prior} or \emph{algorithmic probability}) of a finite
binary string $x$ is:
\[
M(x) \;=\; \sum_{\substack{\pi \in \{0,1\}^*:\\
  U(\pi) \text{ outputs a string}\\ \text{beginning with } x}}
  2^{-|\pi|}
\]
In words: $M(x)$ is the probability that $U$, when given a \emph{random}
program (each bit chosen independently with probability $\frac{1}{2}$),
produces output that starts with $x$.
\end{definition}

\textbf{Prediction via the universal prior.} Given an observed sequence
$x_{1:t} = (x_1, x_2, \ldots, x_t)$, Solomonoff induction predicts the
next symbol $x_{t+1}$ using the conditional distribution derived from $M$:
\[
p_{\text{Sol}}(x_{t+1} \mid x_{1:t})
\;=\; \frac{M(x_{1:t} \cdot x_{t+1})}{M(x_{1:t})}
\]
where $x_{1:t} \cdot x_{t+1}$ denotes the concatenation of $x_{1:t}$ and
$x_{t+1}$, and the ratio is well-defined whenever $M(x_{1:t}) > 0$.
Intuitively, this computes the fraction of ``algorithmic probability
mass'' of programs that, having produced $x_{1:t}$ so far, continue by
outputting $x_{t+1}$.

\textbf{Why this is optimal.} Solomonoff proved a remarkable convergence
result: $M$ \emph{dominates} every computable probability distribution.
Formally, for any computable measure $\mu$ on infinite binary sequences,
there exists a constant $c_\mu > 0$ (depending on $\mu$ but not on $x$)
such that:
\[
M(x) \;\geq\; c_\mu \cdot \mu(x) \qquad \text{for all finite strings } x.
\]
This means that Solomonoff's predictor converges to the true data-generating
distribution $\mu$ at a rate of $O(K(\mu) / t)$ in terms of expected
KL divergence per step, regardless of what $\mu$ is --- as long as $\mu$
is computable. In other words, Solomonoff induction is the best possible
general-purpose predictor: it learns any computable pattern, and it does
so as fast as is theoretically possible (up to the constant $K(\mu)$,
which is the Kolmogorov complexity of the true distribution).

\textbf{Connection to compression.} Solomonoff induction implicitly
performs compression: programs that are short (i.e., that compress the data
well) receive exponentially more weight ($2^{-|\pi|}$). The prediction is
dominated by the simplest (most compressive) explanations consistent with
the data. This is Occam's razor, formalised.

\textbf{Uncomputability.} Like $K(x)$, the Solomonoff prior $M(x)$ is
uncomputable --- computing it would require summing over all programs,
which is equivalent to solving the halting problem. Solomonoff induction
is a theoretical ideal, not a practical algorithm. But it provides the
gold standard against which all practical prediction methods ---
including neural language models --- can be measured.


\subsubsection{AIXI: Universal Artificial Intelligence (Hutter, 2000, 2005)}

Shannon showed that prediction = compression. Solomonoff showed that the
optimal predictor is the one that implicitly finds the shortest programs
(best compressions) consistent with the data. But prediction alone is not
intelligence: intelligence also requires \emph{acting in an environment to
achieve goals}. Marcus Hutter's AIXI framework (2000, monograph 2005)
extends Solomonoff induction from passive prediction to active,
goal-directed agency.

\textbf{The agent--environment setup.} Consider an agent interacting with
an environment over discrete time steps $t = 1, 2, 3, \ldots$ The
interaction proceeds as follows:
\begin{enumerate}[leftmargin=2em]
  \item At each time step $t$, the agent chooses an \textbf{action}
    $a_t \in \mathcal{A}$ from a finite set of possible actions.
  \item The environment responds with an \textbf{observation}
    $o_t \in \mathcal{O}$ and a \textbf{reward}
    $r_t \in \mathcal{R} \subset \mathbb{R}$.
  \item The agent's goal is to choose a sequence of actions
    $a_1, a_2, a_3, \ldots$ that maximises the total expected
    future reward $\sum_{t=1}^{\infty} r_t$ (or a discounted variant).
\end{enumerate}
The environment is modelled as a function (possibly stochastic) that maps
the history of actions and observations so far to the next observation and
reward. The agent does not know the environment's dynamics in advance; it
must learn from interaction.

\textbf{The AIXI agent.} AIXI combines Solomonoff induction (for learning
the environment) with expectimax planning (for choosing optimal actions).
At each time step $t$, having observed history $h_{<t} = (a_1, o_1, r_1,
\ldots, a_{t-1}, o_{t-1}, r_{t-1})$, AIXI selects the action:
\[
a_t^{\text{AIXI}} \;=\; \arg\max_{a_t} \sum_{o_t, r_t}
  \left[\, r_t + \max_{a_{t+1}} \sum_{o_{t+1}, r_{t+1}}
  \left[\, r_{t+1} + \cdots \,\right]
  \cdot M(o_{t:m}, r_{t:m} \mid h_{<t}, a_{t:m}) \,\right]
\]
where $m$ is the planning horizon and $M(\cdot \mid \cdot)$ is the
Solomonoff-style universal mixture over all computable environment
models. In words: AIXI considers all possible futures (all possible
sequences of observations and rewards), weighs them by the universal
prior (which favours simple, compressive environment models), and
chooses the action that maximises expected cumulative reward under this
mixture.

\textbf{Optimality.} Hutter proved that AIXI is \emph{Pareto optimal}
in the following sense: no other agent can achieve higher expected reward
than AIXI in \emph{any} computable environment without achieving lower
expected reward in some other computable environment. In simpler terms,
AIXI is the best possible general-purpose agent --- it learns any
computable environment as fast as theoretically possible and acts
optimally given its knowledge.

\textbf{Why this closes the chain.} AIXI establishes the final link:
\begin{enumerate}[leftmargin=2em]
  \item Shannon (1948): Prediction = Compression.
  \item Solomonoff (1964): Optimal prediction requires optimal compression
    (finding the shortest programs consistent with the data).
  \item Hutter (2000, 2005): Optimal compression, combined with
    expectimax planning, yields the most intelligent possible agent.
    Therefore: Prediction = Compression = Intelligence.
\end{enumerate}
Like Solomonoff induction, AIXI is uncomputable. It is a mathematical
ideal, not a blueprint for building an AI system. But it provides the
theoretical foundation for a striking claim: \emph{any system that
improves at next-token prediction is, in a precise mathematical sense,
becoming more generally intelligent}.

\begin{keyinsight}
The prediction--compression--intelligence chain is now complete:

\begin{center}
\begin{tikzpicture}[
    box/.style={draw, rounded corners=3pt, minimum width=2.6cm,
                minimum height=1cm, align=center, font=\bfseries},
    eq/.style={font=\Large\bfseries, text=accent}
  ]
  \node[box] (pred) {Prediction};
  \node[eq, right=0.4cm of pred] {=};
  \node[box, right=0.8cm of pred] (comp) {Compression};
  \node[eq, right=0.4cm of comp] {=};
  \node[box, right=0.8cm of comp] (intel) {Intelligence};

  \node[below=0.3cm of pred, font=\small, text=muted] {Shannon, 1948};
  \node[below=0.3cm of comp, font=\small, text=muted] {Kolmogorov, 1963};
  \node[below=0.3cm of intel, font=\small, text=muted] {Hutter, 2005};
\end{tikzpicture}
\end{center}

The perfect compressor is the most intelligent possible agent. Not by
analogy, but by mathematical proof. This means that every improvement in
next-token prediction is an improvement in \emph{general intelligence},
in the precise Hutter sense.
\end{keyinsight}

\begin{speakingnote}
For a non-specialist audience, you do not need to present AIXI's
mathematical details. The key message is: ``There is a chain of
mathematical results, published over 60 years, proving that prediction,
compression, and intelligence are the same thing. This is not a metaphor.
It is a theorem.''
\end{speakingnote}


\subsection{The Empirical Bombshell: Language Modeling Is Compression}

D\'{e}letang et al.\ (ICLR 2024, Google DeepMind, arXiv:2309.10668)
performed a deceptively simple experiment: they used language models as
general-purpose compressors and compared them to domain-specific
compression algorithms.

\subsubsection{Arithmetic Coding: Turning Predictions into Compression}
\label{sec:arithmetic}

The key technical idea is \textbf{arithmetic coding}, an algorithm that
converts any probabilistic model into a lossless compressor. We explain it
step by step.

\textbf{Setup.} Suppose we want to encode a sequence $x_1, x_2, \ldots,
x_T$ over alphabet $\mathcal{A}$, and we have a model
$p_\theta$ that, for each position $t$, assigns a conditional distribution
$p_\theta(\cdot \mid x_{<t})$ over the next symbol.

\textbf{The algorithm.} Arithmetic coding works by maintaining a
\emph{subinterval} of $[0, 1)$ that narrows as each symbol is encoded:
\begin{enumerate}[leftmargin=2em]
  \item \textbf{Initialise:} Set the current interval to $[0, 1)$.
  \item \textbf{For each symbol $x_t$ ($t = 1, 2, \ldots, T$):}
    query the model for $p_\theta(\cdot \mid x_{<t})$; partition the
    current interval into sub-intervals with widths proportional to the
    probabilities $p_\theta(a \mid x_{<t})$ for each $a \in \mathcal{A}$;
    then select the sub-interval corresponding to the actual symbol $x_t$.
    This becomes the new current interval.
  \item \textbf{Output:} Any binary fraction within the final interval
    uniquely identifies the encoded sequence. The number of bits needed
    is approximately $-\log_2 w$, where $w$ is the final interval width.
\end{enumerate}

\textbf{Code length analysis.} After encoding all $T$ symbols, the width
of the final interval is:
\[
\text{width} \;=\; \prod_{t=1}^{T} p_\theta(x_t \mid x_{<t})
\]
The number of bits required is:
\[
L(x_{1:T}) \;=\; \lceil -\log_2 \text{width} \rceil + 1
\;=\; \left\lceil \sum_{t=1}^{T} \bigl(-\log_2 p_\theta(x_t \mid
x_{<t})\bigr) \right\rceil + 1
\]
The $+1$ is a technical overhead that becomes negligible for long
sequences. The key insight: \textbf{each symbol $x_t$ contributes
exactly $-\log_2 p_\theta(x_t \mid x_{<t})$ bits to the compressed
length.} If the model predicts $x_t$ with high probability, its
contribution is small (cheap to encode). If the model is surprised by
$x_t$, its contribution is large (expensive to encode).

\textbf{Optimality.} If $p_\theta$ equals the true data distribution
$p^*$, then arithmetic coding achieves the entropy rate --- the
information-theoretically optimal compression. If $p_\theta$ differs from
$p^*$, the code length exceeds the optimum by exactly the
KL divergence:
\[
\frac{1}{T}\sum_{t=1}^{T}\bigl(-\log_2 p_\theta(x_t \mid x_{<t})\bigr)
\;=\; H(X_t \mid X_{<t}) + D_{\text{KL}}(p^* \| p_\theta)
\]
where $H$ is the true conditional entropy and $D_{\text{KL}} \geq 0$ is
the KL divergence between the true and model distributions (averaged over
time steps). A perfect model ($D_{\text{KL}} = 0$) achieves perfect
compression.


\subsubsection{The D\'{e}letang et al.\ Experiment}

\textbf{Method.} D\'{e}letang et al.\ applied arithmetic coding using
Chinchilla 70B (a large language model trained exclusively on text) as
the predictive model $p_\theta$. They fed raw bytes of data from domains
the model had never encountered during training:

\begin{itemize}[leftmargin=2em]
  \item \textbf{Images:} Patches from ImageNet, converted to raw byte
    sequences (pixel values).
  \item \textbf{Audio:} Clips from LibriSpeech, converted to raw byte
    sequences (audio samples).
\end{itemize}

The model processes these bytes exactly as it would process text: one
byte at a time, predicting the next byte given all previous bytes.
The resulting conditional probabilities are fed into an arithmetic coder
to produce the compressed output.

\textbf{Results:}

\begin{center}
\begin{tabular}{lcc}
\toprule
\textbf{Domain} & \textbf{Specialist Compressor} & \textbf{Chinchilla 70B} \\
\midrule
Images (ImageNet)  & PNG: 58.5\% & \textbf{43.4\%} \\
Audio (LibriSpeech) & FLAC: 30.3\% & \textbf{16.4\%} \\
\bottomrule
\end{tabular}
\end{center}
\small{(Compression ratio = compressed size / original size. Lower is better.)}
\normalsize

\begin{keyinsight}
Chinchilla 70B has never seen a photograph. It has never heard a sound.
It was trained on text and text alone. Yet it compresses images better
than PNG (an algorithm \emph{designed} for images) and audio better than
FLAC (an algorithm \emph{designed} for audio).

This is not a parlour trick. It means that next-token prediction on text
forced the model to learn something \emph{universal} about the statistical
structure of data --- structure that transfers across modalities the model
was never exposed to.
\end{keyinsight}

\textbf{Why does this work?} The intuition is that all structured data ---
images, audio, text, code --- shares deep statistical regularities.
Natural images have spatial correlations (nearby pixels tend to have
similar values), frequency structure (smooth regions with sharp edges),
and semantic structure (objects, textures, scenes). Audio has temporal
correlations, harmonic structure, and phonemic patterns. Text
\emph{describes} all of these phenomena. A model that deeply learns the
statistical structure of natural language --- which describes images,
sounds, physical processes, and abstract concepts --- has implicitly
learned regularities shared by the underlying data.

More formally: if the model learns a good approximation to the universal
prior (the Solomonoff distribution), it should compress \emph{any}
computable data well, regardless of domain. The D\'{e}letang result
suggests that large language models are, in fact, approximating something
close to a universal compressor.

\textbf{Connection to scaling laws.} Kaplan et al.\ (2020) showed that
language model loss (= compression rate) follows clean power laws in
model size, dataset size, and compute:
\[
L(N) \;\approx\; \left(\frac{N_c}{N}\right)^{\!\alpha_N}
\]
where $N$ is the number of model parameters, $N_c$ is a fitted constant,
and $\alpha_N \approx 0.076$. These power laws \emph{are} compression
improvement curves: each order of magnitude in scale buys a predictable,
diminishing improvement in compression rate. The scaling laws tell us
that we are on a smooth trajectory toward better and better
approximations of the universal compressor.


\subsection{The Philosophical Punchline of Topic A}

\begin{quote}
\emph{If compression is understanding, and LLMs are the best compressors
we have ever built, then the ``stochastic parrot'' critique is not just
wrong --- it is incoherent.}
\end{quote}

This is a strong claim. Unpack it:

The ``stochastic parrot'' framing (Bender et al., 2021) argues that LLMs
are merely doing sophisticated pattern matching without genuine
understanding. But if the Shannon--Hutter chain is correct, then
``pattern matching'' at the level of a universal compressor \textbf{is}
the most general form of intelligence. There is no ``genuine
understanding'' that is separate from, or above, perfect compression.

\begin{warning}
This conclusion is logically valid \emph{given} the premises. But the
premises are debatable:
\begin{enumerate}[leftmargin=2em]
  \item Hutter's framework defines intelligence in terms of
    compression/prediction. Other definitions exist (embodied cognition,
    phenomenal consciousness, social intelligence).
  \item LLMs are finite approximations to the Solomonoff ideal. The gap
    between the approximation and the ideal may be where ``real''
    understanding lives.
  \item Compression is necessary for intelligence, but it may not be
    \emph{sufficient}. A perfect compressor might still lack something ---
    intentionality, consciousness, grounding.
\end{enumerate}
Do \textbf{not} present the compression = understanding claim as settled
fact. Present it as a serious theoretical argument with strong empirical
support that demands engagement. The audience should leave \emph{unsettled},
not convinced.
\end{warning}

\begin{speakingnote}
The right tone for the transition is: ``I have shown you a mathematical
identity and a shocking empirical result. The implication seems to be that
LLMs genuinely understand. But I want you to hold that conclusion lightly,
because I am about to show you something that makes it even stranger ---
and then something that complicates it.''
\end{speakingnote}


% ============================================================================
\section{Topic B: The Platonic Representation Hypothesis}
% ============================================================================

\subsection{Opening: Plato's Allegory of the Cave}

Before presenting any empirical results, set the philosophical frame.

Plato's \emph{Republic}, Book VII (c.\ 375 BCE): Prisoners chained in a
cave see only shadows on the wall. They take the shadows for reality. One
prisoner escapes, sees the sun and real objects, and realises the shadows
were mere projections of a deeper reality.

Plato's metaphysics posits a ``realm of Forms'' --- perfect, abstract
templates of which physical objects are imperfect copies. A chair is a
chair because it participates in the Form of Chair-ness. This view has
been debated for 2,400 years.

\begin{speakingnote}
``Plato argued 2,400 years ago that behind the messy world of appearances
lies a realm of perfect Forms. Philosophers have been arguing about it
ever since. In 2024, a team at MIT published evidence that AI models are
converging on the same internal representation of reality --- regardless
of who built them, what data they trained on, or what architecture they
used. Plato, it turns out, may have been writing about neural networks.''
\end{speakingnote}


\subsection{The Hypothesis (Huh et al., ICML 2024)}

Huh, Cheung, Wang, and Isola (MIT) published ``The Platonic Representation
Hypothesis'' as a position paper at ICML 2024 (arXiv:2405.07987). Their
central claim:

\begin{keyinsight}[title={The Platonic Representation Hypothesis}]
As AI models increase in scale and capability, their learned
representations converge toward a shared statistical model of reality,
regardless of:
\begin{itemize}[leftmargin=2em]
  \item The training data (text, images, audio, video, code)
  \item The architecture (Transformer, CNN, diffusion model, etc.)
  \item The training objective (next-token prediction, contrastive
    learning, masked image modelling, etc.)
  \item The organisation that built the model
\end{itemize}
The hypothesis is that there exists a unique ``correct'' representation
of reality's statistical structure, and all sufficiently powerful learners
discover it.
\end{keyinsight}

\subsection{How the Evidence Is Measured}

Huh et al.\ use two main quantitative tools to measure whether different
models' internal representations are converging: \textbf{Centred Kernel
Alignment (CKA)} and \textbf{mutual nearest neighbours (MNN)}. We explain
each.

\subsubsection{Centred Kernel Alignment (CKA)}

\begin{definition}[title={Centred Kernel Alignment}]
Let $\mathbf{X} \in \mathbb{R}^{n \times p}$ and $\mathbf{Y} \in
\mathbb{R}^{n \times q}$ be two \textbf{representation matrices}, where
$n$ is the number of data points (e.g., images or sentences), $p$ is the
representation dimension of model A, and $q$ is the representation
dimension of model B. Each row of $\mathbf{X}$ is model A's internal
representation (e.g., the hidden-state vector at a particular layer) of
the $i$-th data point; similarly for $\mathbf{Y}$ and model B.

The (linear) \textbf{CKA} between these representations is defined as:
\[
\operatorname{CKA}(\mathbf{X}, \mathbf{Y})
\;=\; \frac{\|\mathbf{Y}^\top \mathbf{X}\|_F^2}
  {\|\mathbf{X}^\top \mathbf{X}\|_F \;\cdot\;
   \|\mathbf{Y}^\top \mathbf{Y}\|_F}
\]
where $\|\cdot\|_F$ denotes the Frobenius norm:
$\|\mathbf{A}\|_F = \sqrt{\sum_{i,j} A_{ij}^2}$.
\end{definition}

\textbf{Interpretation.}
\begin{itemize}[leftmargin=2em]
  \item $\operatorname{CKA} = 1$ means the two representations encode
    the same information about the data (up to a linear transformation).
  \item $\operatorname{CKA} = 0$ means the representations are completely
    unrelated.
  \item CKA is \textbf{invariant to orthogonal transformations and
    isotropic scaling}: if you rotate model B's representation space or
    scale it uniformly, CKA does not change. This makes it appropriate
    for comparing models with different architectures and dimensionalities,
    because a representation can be ``the same'' even if it is expressed in
    a different coordinate system.
\end{itemize}

\textbf{How Huh et al.\ use it.} For each pair of models (e.g., a vision
Transformer and a language model), they feed the same set of $n$ inputs
through both models, extract internal representations (hidden states at a
particular layer), form the matrices $\mathbf{X}$ and $\mathbf{Y}$, and
compute CKA. They repeat this for models of varying sizes and observe
that CKA \emph{increases} as both models get larger. Small models have
idiosyncratic representations; large models converge.


\subsubsection{Mutual Nearest Neighbours (MNN)}

CKA measures global, linear similarity between representation spaces. MNN
captures a complementary, local notion of alignment.

\begin{definition}[title={Mutual Nearest Neighbours}]
Given two sets of representations for the same $n$ data points ---
$\{\mathbf{x}_1, \ldots, \mathbf{x}_n\}$ from model A and
$\{\mathbf{y}_1, \ldots, \mathbf{y}_n\}$ from model B --- the
\textbf{mutual nearest-neighbour score} is computed as follows:
\begin{enumerate}[leftmargin=1.5em]
  \item For each data point $i$, find its $k$ nearest neighbours in model
    A's representation space. Call this set $\mathcal{N}_A(i)$.
  \item Similarly, find its $k$ nearest neighbours in model B's space:
    $\mathcal{N}_B(i)$.
  \item Compute the overlap: $\operatorname{MNN}_k(i) =
    |\mathcal{N}_A(i) \cap \mathcal{N}_B(i)| / k$.
  \item Average over all data points:
    $\operatorname{MNN}_k = \frac{1}{n}\sum_{i=1}^{n}
    \operatorname{MNN}_k(i)$.
\end{enumerate}
Here ``nearest neighbours'' are determined by cosine similarity or
Euclidean distance in the respective representation spaces. The parameter
$k$ is typically set to a small value (e.g., $k = 10$).
\end{definition}

\textbf{Interpretation.} If two models organise their representations
similarly, then data points that are ``close together'' in one model's
space should also be close together in the other's. A high MNN score
means the local neighbourhood structure is preserved across models.
A low MNN score (close to $k/n$, the chance level) means the
representations organise data in fundamentally different ways.

\textbf{Why MNN matters.} CKA captures global, linear structure, but can
miss local differences (e.g., two representations might have the same
overall covariance structure but differ in fine-grained neighbourhood
structure). MNN captures exactly this fine-grained structure. The fact
that \emph{both} CKA and MNN increase with scale strengthens the
convergence claim.


\subsubsection{Cross-Model and Cross-Modality Evidence}

Huh et al.\ report three key findings, all measured using CKA and MNN:

\begin{enumerate}[leftmargin=2em]
  \item \textbf{Cross-model alignment increases with scale.} Pairs of
    vision models (e.g., different ResNets, Vision Transformers, DINO
    models) show increasing CKA and MNN as both models are scaled up.
    The same holds for pairs of language models.

  \item \textbf{Cross-modality alignment increases with scale.} Even
    vision models and language models --- processing entirely different
    data types --- develop increasingly aligned representation spaces.
    To measure this, Huh et al.\ use paired datasets where each data
    point has both a visual and a textual representation (e.g., an image
    and its caption). They extract model A's representation of the image
    and model B's representation of the caption, and compute CKA/MNN.
    The alignment increases with scale.

  \item \textbf{Convergence toward a common point.} The pattern is not
    just pairwise: multiple models appear to converge toward a single
    shared representation, suggesting a unique attractor in
    representation space.
\end{enumerate}


\subsection{The Kernel Hypothesis}

Huh et al.\ propose a theoretical explanation for why convergence occurs.
Their argument, which they call the \textbf{kernel hypothesis}, proceeds
as follows.

\textbf{Setup.} Suppose reality can be described by a joint distribution
over all possible sensory modalities:
\[
p^*(\text{text}, \text{image}, \text{audio}, \text{video}, \ldots)
\]
This is the ``true'' joint distribution over everything that could be
observed. No single model sees all modalities; each model is trained on a
\emph{marginal} (a subset of modalities). A language model sees the
text marginal $p^*(\text{text})$; a vision model sees the image marginal
$p^*(\text{image})$.

\textbf{The kernel.} Different modalities are not independent --- they are
all generated by the same underlying reality. For instance, the visual
appearance of a cat, the word ``cat,'' the sound of a meow, and a video
of a cat playing are all \emph{different projections of the same
underlying entity}. The \textbf{kernel} of the joint distribution is the
shared structure that is common to all modalities --- informally, it is
the ``shape of reality'' that remains after you strip away
modality-specific details.

More formally, consider the pointwise mutual information (PMI) matrix
between items across modalities. The \emph{kernel} of this matrix ---
the subspace spanned by its dominant eigenvectors --- captures the shared
statistical structure. Huh et al.\ argue that as models become more
powerful, they learn to represent this kernel, because the kernel captures
the regularities that are most useful for prediction \emph{regardless of
modality}.

\textbf{Why convergence occurs.} A sufficiently powerful model, trained
on \emph{any} modality, will eventually discover the kernel of the joint
distribution, because that kernel is the structure most useful for
prediction (and therefore compression). Since the kernel is unique (a
property of reality, not of the model), all sufficiently powerful models
converge on the same representation.

\textbf{Analogy.} Imagine three artists --- one blind, one deaf, one
limited to touch --- each trying to represent a sculpture. The blind
artist works from descriptions; the deaf artist from photographs; the
touch-limited artist from physical contact. If they are skilled enough,
their representations will converge on the same underlying 3D shape,
because that shape is the common cause of all their sensory data. The
kernel hypothesis says AI models are doing the same thing.


\subsection{Evidence: Othello-GPT and Emergent World Models}

Li et al., ``Emergent World Representations: Exploring a Sequence Model
Trained on a Synthetic Task'' (ICLR 2023).

\textbf{Setup.} A GPT-style Transformer (8 layers, 8 attention heads,
embedding dimension 512) is trained on sequences of legal Othello moves.
Each move is represented as a single token indicating the board position
(e.g., token ``d3'' means ``place a piece at column d, row 3''). The
model receives:
\begin{itemize}[leftmargin=2em]
  \item No description of the Othello board (no $8 \times 8$ grid)
  \item No encoding of the rules (no information about captures or flips)
  \item No visual representation of game states
  \item Only raw move sequences from recorded games
\end{itemize}
The training objective is standard next-token prediction: given the moves
so far, predict the next move.

\textbf{Finding.} Li et al.\ trained \textbf{non-linear probes}
(small neural networks) on the model's internal activations to predict
the board state --- which of the 64 cells are black, white, or empty ---
at each point in the game. The probes achieved high accuracy, indicating
that the model's internal representations encode the board state, even
though the board was never part of the training data.


\subsubsection{The Nanda Refinement: Linear Probes}

Neel Nanda (``Actually, Othello-GPT Has A Linear Emergent World
Representation,'' 2023) strengthened this finding substantially.

\textbf{The probe methodology.} A \textbf{probe} is a small, separately
trained model that takes a neural network's internal activations as input
and tries to predict some property of interest (here, the board state).
The key question is: how complex does the probe need to be?

\begin{itemize}[leftmargin=2em]
  \item A \textbf{non-linear probe} (e.g., a small multi-layer network)
    can extract information even if it is encoded in a complex, tangled
    way in the activations. High probe accuracy with a non-linear probe
    shows the information is \emph{present}, but does not reveal how it
    is organised.
  \item A \textbf{linear probe} is a single matrix multiplication:
    $\hat{y} = \mathbf{W} \mathbf{h} + \mathbf{b}$, where $\mathbf{h}
    \in \mathbb{R}^d$ is the model's hidden-state vector at a given layer,
    $\mathbf{W} \in \mathbb{R}^{c \times d}$ is a learned weight matrix,
    $\mathbf{b} \in \mathbb{R}^c$ is a bias vector, and $c$ is the number
    of classes (here $c = 3$: black, white, empty, for each cell). High
    accuracy with a linear probe means the information is encoded
    \emph{linearly} --- as a direction in the representation space.
\end{itemize}

\textbf{Nanda's result.} Linear probes on Othello-GPT's middle-layer
activations achieve high accuracy (over 95\%) in classifying each cell as
black, white, or empty. This means the board state is not just present
somewhere in the model's representations --- it is encoded as a
\emph{linear} function of the hidden-state vector. Each cell's state
corresponds to a direction in the 512-dimensional representation space.

\textbf{Why linearity matters.}
\begin{enumerate}[leftmargin=2em]
  \item \textbf{Interpretability:} A linear representation means the
    board state is geometrically ``clean'' --- each cell's state
    corresponds to a hyperplane in representation space, and the states
    can be read off by projection.
  \item \textbf{Not memorisation:} A linear probe is too simple to
    memorise complex mappings. If a linear probe succeeds, the
    information must be explicitly and systematically encoded in the
    activations, not recoverable only through a complex non-linear
    computation.
  \item \textbf{Compositionality:} Linear representations compose
    naturally (sums of directions remain meaningful), which suggests the
    model has learned a structured, modular encoding of the board state
    rather than an opaque blob of features.
\end{enumerate}

\begin{keyinsight}
The model \emph{inferred the existence of a world it was never shown}.
It was trained to predict the next move in a sequence. From that
prediction task alone, it constructed an internal, geometrically clean,
linear model of the hidden state that generates the sequence. This is
exactly what the prediction--compression--intelligence chain predicts:
to compress move sequences optimally, you need to discover the latent
structure (the board) that generates them.
\end{keyinsight}

\begin{speakingnote}
Emphasise the wonder here: ``A model trained on sequences of Othello
moves --- just coordinate pairs --- was never told that a board exists,
or what the rules are. Inside its weights, researchers found a clean,
linear representation of the board state. The model inferred the existence
of a world it was never shown. If that doesn't make you pause, I don't
know what will.''
\end{speakingnote}


\subsection{Evidence: Space and Time Neurons}

Gurnee and Tegmark, ``Language Models Represent Space and Time'' (ICLR 2024).

\textbf{Setup.} The authors probed the internal activations of Llama-2
(a large language model trained on text) when it processed sentences
containing geographic locations (cities, countries) and historical dates.
They used linear probes (as described above) to test whether individual
neurons or linear directions in the representation space encode spatial
coordinates or temporal information.

\textbf{Findings:}

\begin{enumerate}[leftmargin=2em]
  \item \textbf{Space neurons:} Individual neurons in Llama-2 activate
    in proportion to the latitude or longitude of the location mentioned
    in the input text. A linear regression from a single neuron's
    activation to the geographic coordinate of the mentioned city yields
    high $R^2$ values. The model has built an internal coordinate system
    --- essentially, a map --- from text alone.

  \item \textbf{Time neurons:} Similarly, individual neurons encode
    historical dates on a linear timeline. The activation is proportional
    to the year associated with the entity or event in the text.

  \item \textbf{Linearity:} Both spatial and temporal representations are
    \emph{linear} --- they can be extracted by simple linear probes (a
    matrix multiplication from the model's hidden state to a predicted
    coordinate). This mirrors the Othello-GPT finding and suggests a
    general principle: language models build clean, structured, linear
    models of latent variables that govern their training data.
\end{enumerate}

\textbf{Interpretation.} The model was never shown a map, a globe, or a
timeline. It learned spatial and temporal structure purely from the
statistical regularities of text --- which cities co-occur in sentences,
which events are described as temporally adjacent, how weather and
cultural descriptions cluster by region. The model discovered that the
latent generating structure behind these textual patterns is a spatial
coordinate system and a temporal axis.

This connects directly to the Platonic Representation Hypothesis: the
model is recovering the \emph{actual structure of reality} (geography,
history) from a lossy projection of that structure (text). Different
models trained on different text corpora converge on similar spatial and
temporal representations, just as the hypothesis predicts.


\subsection{Business Implications: The Commodity Question}

If the Platonic Representation Hypothesis is correct, it has a profound
implication for AI strategy:

\begin{quote}
\emph{If all foundation models converge on the same representation of
reality, does it matter which one you buy? Or is intelligence a commodity
--- like electricity --- where the only question is price?}
\end{quote}

This is deliberately provocative. The nuanced answer:

\begin{enumerate}[leftmargin=2em]
  \item \textbf{We are not there yet.} Current models still differ
    meaningfully in capabilities, safety properties, and domain
    performance. The convergence is a trend, not a fact.

  \item \textbf{Convergence of representations $\neq$ convergence of
    behaviour.} Even if internal representations converge, the mapping
    from representations to outputs (the ``head'' of the model, RLHF
    tuning, system prompts, safety filters) can still differentiate
    products significantly.

  \item \textbf{But the trend matters for long-term strategy.} If the
    underlying intelligence becomes commoditised, then durable competitive
    advantage shifts to data moats, distribution, user experience,
    domain-specific fine-tuning, integration ecosystems, and trust/safety
    --- not raw model capability. Companies building AI strategies should
    not bet exclusively on one model provider's unique capabilities.

  \item \textbf{Invest in portability.} Use abstraction layers (e.g.,
    standardised API interfaces, prompt management frameworks) that allow
    you to switch model providers without rebuilding your stack.
\end{enumerate}

\begin{speakingnote}
This is where you connect the philosophy to the audience's actual
decisions. They are making model selection and vendor lock-in decisions
right now. The Platonic Representation Hypothesis suggests that betting
on the long-term superiority of any single model provider is risky.
\end{speakingnote}


\subsection{Caveats and Criticisms}

Be honest about the limitations of the Platonic Representation Hypothesis:

\begin{enumerate}[leftmargin=2em]
  \item \textbf{It is a position paper, not a proof.} The evidence is
    suggestive, not conclusive. The trend of increasing alignment could
    plateau or reverse.

  \item \textbf{``Convergence'' is measured with specific metrics} (CKA,
    mutual nearest neighbours) that capture particular notions of
    similarity. Other metrics might show different results. CKA captures
    linear, global structure; MNN captures local neighbourhood structure;
    but there may be aspects of representation quality that neither
    metric captures.

  \item \textbf{The claim is about representations, not capabilities.}
    Two models could have similar representations but very different
    abilities to \emph{use} those representations for downstream tasks.

  \item \textbf{Selection bias.} Most large-scale models are Transformers
    trained on internet data. The convergence might reflect convergence of
    architectures and training data, not a universal property of
    intelligence.

  \item \textbf{The ``unique representation'' claim is metaphysical.}
    Saying there is ``one correct'' representation of reality is a strong
    philosophical commitment (scientific realism of a specific kind).
    Anti-realists and pragmatists would push back.
\end{enumerate}


% ============================================================================
\section{Topic C: The Chinese Room, 45 Years Later}
% ============================================================================

\subsection{Searle's Original Argument (1980)}

John Searle, ``Minds, Brains, and Programs,'' \emph{Behavioral and Brain
Sciences}, 1980.

\textbf{The thought experiment:}

\begin{enumerate}[leftmargin=2em]
  \item A person (who speaks only English) is locked in a room.
  \item Chinese characters are passed through an input slot.
  \item The person consults an enormous rulebook that specifies: ``If you
    receive characters $X$, output characters $Y$.''
  \item The person follows the rules mechanically and passes Chinese
    characters back through an output slot.
  \item To an observer outside the room, the room speaks perfect Chinese.
  \item But the person inside understands \emph{no Chinese whatsoever}.
\end{enumerate}

\textbf{Searle's conclusion:} The room implements a program that passes
the Turing test for Chinese, but there is no understanding of Chinese
anywhere in the system. Therefore: \textbf{syntax is not sufficient for
semantics.} Computation (symbol manipulation according to rules) cannot,
by itself, produce genuine understanding. No matter how complex the
program, no matter how perfect the output, something essential is missing.

\textbf{Searle's target:} ``Strong AI'' --- the claim that a suitably
programmed computer literally \emph{has} a mind, understands, and has
cognitive states. Searle accepts ``weak AI'' (computers as useful tools
for studying the mind) but rejects strong AI.

\subsection{The Classical Responses}

Searle's paper provoked immediate and sustained debate. The major
counter-arguments:

\begin{enumerate}[leftmargin=2em]
  \item \textbf{The Systems Reply:} The person in the room doesn't
    understand Chinese, but the \emph{system as a whole} (person +
    rulebook + room) does. Searle's response: imagine the person
    memorises the entire rulebook and does the work in their head. They
    still don't understand Chinese. (But this response is itself debated
    --- does internalising the program change the situation?)

  \item \textbf{The Robot Reply:} Understanding requires a body that
    interacts with the world. Give the system sensors and actuators, and
    it might understand. Searle's response: this merely adds more
    syntactic processing; you can imagine a robot whose entire control
    system is the Chinese Room.

  \item \textbf{The Brain Simulator Reply:} What if the program simulates
    a Chinese speaker's brain at the neuron level? Searle's response:
    even then, the simulation is just symbol manipulation. Simulating
    digestion doesn't produce actual digestion.

  \item \textbf{The Other Minds Reply:} We never have direct access to
    anyone else's understanding. We infer understanding from behaviour.
    If the room's behaviour is indistinguishable from understanding, on
    what basis do we deny it? Searle's response: this is behaviourism,
    which he rejects.
\end{enumerate}

\begin{speakingnote}
You do not need to go through all four replies in the presentation ---
time is short. Mention the systems reply and the other minds reply, as
they are the most relevant to the LLM case. The key point is that
\emph{none} of these replies is decisive. After 45 years, the debate
is not settled.
\end{speakingnote}


\subsection{Updating the Chinese Room for LLMs}

Now retell Searle's thought experiment with modern parameters:

\begin{quote}
``The rulebook is 175 billion parameters. The `symbols' are every sentence
humans have ever published on the internet. The `room' processes inputs
at superhuman speed and produces outputs that fool domain experts.''
\end{quote}

The question is: does scaling the Chinese Room change the argument? Three
bodies of evidence bear on this question.

\subsubsection{Evidence For Understanding}

\textbf{1. Emergent world models (Othello-GPT, Li et al., ICLR 2023).}
The model doesn't just shuffle symbols --- it builds an internal
representation of the hidden state that generates those symbols. This is
harder to dismiss as ``mere symbol manipulation'' because the world model
was never present in the training data. The model \emph{inferred} it.

\textbf{2. Cross-domain compression (D\'{e}letang et al., ICLR 2024).}
The model compresses data from domains it has never seen. This suggests
it has learned something about the \emph{structure of data itself}, not
just the surface statistics of its training distribution. A compressor
that generalises across domains looks less like a lookup table and more
like a general-purpose understanding engine.

\textbf{3. Spatial and temporal representations (Gurnee \& Tegmark, ICLR
2024).} The model doesn't just process words about places and times ---
it builds geometric representations that mirror the actual structure of
space and time. This is representation, not regurgitation.

Each of these findings undermines a key assumption in Searle's argument:
that what happens inside the room is ``mere syntax'' with no connection
to the structure of the world.


\subsubsection{Evidence Against Understanding}

\textbf{1. The sensorimotor gap (Nature Human Behaviour, 2025).}

The paper ``Large language models without grounding recover non-sensorimotor
but not sensorimotor features of human concepts'' provides the most
nuanced empirical answer to date. The study compared the conceptual
representations of LLMs with those of humans, feature by feature.

\emph{Key findings:}
\begin{itemize}[leftmargin=2em]
  \item LLMs successfully recover \textbf{abstract, linguistic,
    taxonomic} features of concepts: ``a lemon is a citrus fruit,''
    ``lemons are yellow,'' ``lemons are used in cooking.''
  \item LLMs \textbf{fail} on \textbf{sensorimotor and embodied}
    features: what it feels like to hold a lemon, the shock of biting
    into one, the specific quality of its smell, the weight and texture
    in your hand.
  \item This gap persists across model sizes and architectures. It
    appears to reflect a fundamental limitation: the model has never had
    a body, never interacted with a physical lemon, never had sensory
    experience. No amount of text can fully substitute for embodied
    interaction.
\end{itemize}

\begin{keyinsight}
LLMs know that lemons are yellow and sour. They do not know what it
\emph{feels like} to bite one. They have partial understanding ---
the abstract, propositional kind --- but lack the embodied,
experiential kind. This is not a total absence of understanding, nor
a full presence. It is something in between, and our vocabulary does
not have a good word for it.
\end{keyinsight}

\textbf{2. The Vector Grounding Problem (arXiv:2304.01481, 2023, updated
2025).}

This paper argues that multimodality --- adding vision, audio, or
embodiment --- is neither necessary nor sufficient for genuine
grounding. A model can process images and still fail to ``understand''
them in any deep sense. Conversely, some forms of abstract understanding
may be achievable from text alone. The paper shows that the grounding
problem is \emph{harder} than the multimodality debate suggests: adding
sensory data does not automatically solve the symbol grounding problem.

\textbf{3. Contemporary philosophical analysis (Inquiry, 2025).}

``LLMs, Turing Tests and Chinese Rooms'' argues that the field is still
struggling to move beyond Searle's framing. Neither side has a decisive
argument. The paper documents the ongoing philosophical impasse: LLMs
pass increasingly sophisticated versions of the Turing test, but passing
the test was never intended to be \emph{sufficient} for understanding
(even Turing was careful about this). The Chinese Room argument shows that
passing the test is compatible with zero understanding. Yet the empirical
evidence from emergent world models and cross-domain compression suggests
that what is happening inside LLMs is not ``zero understanding'' in any
straightforward sense.


\subsection{The Honest Conclusion}

\begin{tcolorbox}[colback=warnbg, colframe=Red, fonttitle=\bfseries,
  title={The Current State of Knowledge}]
We cannot say whether LLMs understand. Not because we lack data --- we
have more relevant empirical evidence than at any point in the history
of the philosophy of mind. We cannot answer the question because we
\textbf{do not possess a rigorous, agreed-upon definition of
``understanding''} that would let us adjudicate the evidence.

This is not evasion. It is not false modesty. It is the actual state of
knowledge in 2026, acknowledged by leading AI researchers and
philosophers of mind.
\end{tcolorbox}

\textbf{Why this matters practically:}

The absence of a definition is not merely an academic curiosity. It has
direct consequences for every organisation deploying LLMs:

\begin{enumerate}[leftmargin=2em]
  \item \textbf{Trust calibration.} When you deploy an LLM for medical
    triage, legal review, or financial analysis, you are making an
    implicit bet about whether the model ``understands'' the domain
    well enough to be trusted. Without a definition of understanding,
    you cannot rigorously assess this.

  \item \textbf{Regulation.} Legislators are drafting AI regulations that
    depend on concepts like ``understanding,'' ``reasoning,'' and
    ``knowledge.'' These concepts are not well-defined. Regulatory
    frameworks built on undefined terms will be inconsistent and fragile.

  \item \textbf{Liability.} If an LLM causes harm, liability may turn on
    whether the system ``understood'' the risks. Without a definition,
    this is undecidable.

  \item \textbf{Competitive strategy.} If your competitor claims their
    model ``understands'' your industry, do you believe them? On what
    basis? The lack of a shared definition means AI marketing operates
    in a conceptual vacuum.
\end{enumerate}

\begin{speakingnote}
The closing should be slightly unsettling. The audience came in thinking
``understanding'' was a clear concept. They should leave realising it is
not --- and that this conceptual gap has real consequences for their work.
Say something like: ``If we cannot agree on what `understanding' means for
a machine, how should your company decide whether to trust one with
critical decisions? You are already making that bet --- perhaps without
realising that philosophy is the bottleneck.''
\end{speakingnote}


% ============================================================================
\section{Closing and Discussion}
% ============================================================================

\subsection{The Closing Statement}

\begin{quote}
\emph{``The question is not whether machines think. The question is
whether we know what \textbf{we} mean by `thinking.'\,''}
\end{quote}

This line summarises the entire session. The deepest obstacle to answering
questions about machine intelligence is not technical --- it is
conceptual. We do not have adequate definitions of understanding,
knowledge, or intelligence. LLMs have not solved this problem; they have
made it urgent.


\subsection{Discussion Questions}

Use these three questions for the 5-minute discussion. They are ordered
from philosophical to practical:

\begin{enumerate}[leftmargin=2em]
  \item \textbf{If compression is understanding, and LLMs are the best
    compressors we have ever built, should we grant them some form of
    epistemic authority --- or does something essential about
    understanding still require a body, a history, and stakes in the
    outcome?}

    \emph{This gets at the embodiment question. The Nature 2025 result
    suggests embodiment matters for some kinds of understanding but not
    others.}

  \item \textbf{The Platonic Representation Hypothesis implies that
    intelligence may be a commodity. If that is true, what is your
    organisation's durable competitive advantage in a world where the
    ``thinking'' is cheap and interchangeable?}

    \emph{This is the business strategy question. It should provoke the
    executives in the room.}

  \item \textbf{You are already deploying AI systems in contexts where
    the question ``does it understand?'' has real consequences. If
    philosophy has not yet answered that question, what standard of
    evidence should your company use in the meantime? Who is responsible
    for setting that standard?}

    \emph{This is the actionable question. It forces the audience to
    confront the gap between philosophical uncertainty and operational
    necessity.}
\end{enumerate}


% ============================================================================
\section{Technical Appendix: Detailed References and Proofs}
% ============================================================================

This appendix provides additional technical detail for readers who want
to go deeper into the mathematical and philosophical foundations.

\subsection{The Shannon Source Coding Theorem}

\textbf{Theorem} (Shannon, 1948). \emph{Let $X_1, X_2, \ldots$ be a
stationary ergodic source with entropy rate
$H = \lim_{T \to \infty} \frac{1}{T} H(X_1, \ldots, X_T)$.
No lossless compression scheme can achieve an expected code rate below
$H$ bits per symbol. Conversely, there exist codes (e.g., Huffman codes,
arithmetic codes) that achieve rates arbitrarily close to $H$.}

The connection to prediction: for a conditional model
$p_\theta(x_t \mid x_{<t})$, the expected code length per symbol under
arithmetic coding is:
\[
\bar{\ell}(\theta) \;=\;
\mathbb{E}\!\left[-\log_2 p_\theta(X_t \mid X_{<t})\right]
\;=\; H(X_t \mid X_{<t}) \;+\;
D_{\text{KL}}\!\left(p^*(\cdot \mid X_{<t}) \;\|\;
p_\theta(\cdot \mid X_{<t})\right)
\]
where $p^*(\cdot \mid x_{<t})$ is the true conditional distribution of
$X_t$ given the history, $H(X_t \mid X_{<t})$ is the true conditional
entropy, and $D_{\text{KL}}(p^* \| p_\theta)$ is the expected
KL divergence between the true and model distributions (the expectation
is over histories). Since $D_{\text{KL}} \geq 0$, we always have
$\bar{\ell}(\theta) \geq H(X_t \mid X_{<t})$, with equality if and only
if $p_\theta = p^*$ almost surely.


\subsection{Kolmogorov Complexity and Incompressibility}

\textbf{Theorem} (Kolmogorov). $K(x)$ is not computable: there is no
algorithm that, given a string $x$, returns $K(x)$.

\textbf{Proof sketch.} Suppose for contradiction that $K$ is computable.
Define the function $f(n) = $ ``the first string $x$ (in lexicographic
order) with $K(x) > n$.'' Such a string exists for every $n$ because
only finitely many strings have $K(x) \leq n$ (there are at most
$2^{n+1} - 1$ programs of length $\leq n$). The function $f$ is
computable (by assumption), and the program that computes $f(n)$ has
length $O(\log n)$ bits (it needs to encode the constant-size algorithm
plus the value of $n$, which takes $\lceil \log_2(n+1) \rceil$ bits).
But $f(n)$ outputs a string $x$ with $K(x) > n$. Since $f(n)$ is
computed by a program of length $O(\log n)$, we have
$K(x) \leq O(\log n) < n$ for sufficiently large $n$. Contradiction.

\textbf{Relevance:} This is why we use \emph{approximations} to
Kolmogorov complexity. Language models, gzip, PNG, and FLAC are all
compressors that approximate $K(x)$ from above. The D\'{e}letang result
shows that language models are better approximations than domain-specific
algorithms, at least for the tested domains.


\subsection{Solomonoff's Universal Prior (Detailed)}

The Solomonoff prior $M(x)$ (defined in Section 2) satisfies the
following key property:

\textbf{Theorem} (Solomonoff, 1964; Zvonkin \& Levin, 1970).
\emph{For any computable probability measure $\mu$ on $\{0,1\}^\infty$,}
\[
\sum_{t=1}^{\infty} \mathbb{E}_\mu\!\left[
  D_{\text{KL}}\!\left(\mu(\cdot \mid x_{<t}) \;\|\;
  M(\cdot \mid x_{<t})\right)
\right] \;\leq\; K(\mu) \cdot \ln 2
\]
\emph{where $K(\mu)$ is the Kolmogorov complexity of $\mu$ (the length
of the shortest program that computes $\mu$).}

In words: the total expected KL divergence between the true distribution
$\mu$ and Solomonoff's predictor is bounded by $K(\mu) \ln 2$, which is
a finite constant. This means the per-step divergence
$D_{\text{KL}}(\mu \| M)$ converges to zero at a rate of
$O(K(\mu)/t)$, and convergence is faster when the true distribution
is simple (low $K(\mu)$). This is a formal sense in which Solomonoff
induction is optimally efficient: it learns any computable pattern as
fast as is theoretically possible.


\subsection{CKA: Measuring Representation Similarity (Detailed)}

Centred Kernel Alignment was introduced by Kornblith et al.\ (2019) and
adopted by Huh et al.\ for measuring representation convergence. The
linear CKA formula given in the main text can be understood as a
normalised version of the Hilbert--Schmidt Independence Criterion (HSIC):
\[
\operatorname{HSIC}(\mathbf{X}, \mathbf{Y})
\;=\; \frac{1}{(n-1)^2} \operatorname{tr}\!\left(
  \mathbf{X}\mathbf{X}^\top \mathbf{H} \mathbf{Y}\mathbf{Y}^\top
  \mathbf{H}\right)
\]
where $\mathbf{H} = \mathbf{I}_n - \frac{1}{n}\mathbf{1}\mathbf{1}^\top$
is the centering matrix. CKA normalises this:
\[
\operatorname{CKA}(\mathbf{X}, \mathbf{Y})
\;=\; \frac{\operatorname{HSIC}(\mathbf{X}, \mathbf{Y})}
  {\sqrt{\operatorname{HSIC}(\mathbf{X}, \mathbf{X}) \cdot
   \operatorname{HSIC}(\mathbf{Y}, \mathbf{Y})}}
\]
which is equivalent to the formula in the main text (under the linear
kernel, after centering). CKA is invariant to orthogonal transformations
and isotropic scaling of each representation, but \emph{not} invariant to
arbitrary invertible linear transformations --- a deliberate design choice,
since arbitrary linear transformations could trivially make any two
representations ``look the same.''


% ============================================================================
\section{Complete Reference List}
% ============================================================================

\begin{enumerate}[leftmargin=2em, label={[\arabic*]}]
  \item C.~E. Shannon, ``A Mathematical Theory of Communication,''
    \emph{Bell System Technical Journal}, vol.~27, pp.~379--423, 1948.

  \item A.~N. Kolmogorov, ``On Tables of Random Numbers,''
    \emph{Sankhy\=a}, Series A, vol.~25, pp.~369--376, 1963.

  \item R.~J. Solomonoff, ``A Formal Theory of Inductive Inference,''
    \emph{Information and Control}, vol.~7, pp.~1--22, 224--254, 1964.

  \item M.~Hutter, \emph{Universal Artificial Intelligence: Sequential
    Decisions Based on Algorithmic Probability}, Springer, 2005.

  \item G.~D\'{e}letang, A.~Ruoss, P.-A.~Duquenne et al., ``Language
    Modeling Is Compression,'' \emph{Proc.\ ICLR}, 2024.
    arXiv:2309.10668.

  \item J.~Kaplan, S.~McCandlish, T.~Henighan et al., ``Scaling Laws
    for Neural Language Models,'' arXiv:2001.08361, 2020.

  \item M.~Huh, B.~Cheung, T.~Wang, and P.~Isola, ``Position: The
    Platonic Representation Hypothesis,'' \emph{Proc.\ ICML}, 2024.
    arXiv:2405.07987.

  \item K.~Li, A.~K.~Hopkins, D.~Bau et al., ``Emergent World
    Representations: Exploring a Sequence Model Trained on a Synthetic
    Task,'' \emph{Proc.\ ICLR}, 2023.

  \item N.~Nanda, ``Actually, Othello-GPT Has A Linear Emergent World
    Representation,'' Blog post, 2023.

  \item W.~Gurnee and M.~Tegmark, ``Language Models Represent Space and
    Time,'' \emph{Proc.\ ICLR}, 2024.

  \item J.~R. Searle, ``Minds, Brains, and Programs,'' \emph{Behavioral
    and Brain Sciences}, vol.~3, no.~3, pp.~417--424, 1980.

  \item E.~M. Bender, T.~Gebru, A.~McMillan-Major, and S.~Shmitchell,
    ``On the Dangers of Stochastic Parrots: Can Language Models Be Too
    Big?'' \emph{Proc.\ FAccT}, 2021.

  \item ``Large language models without grounding recover
    non-sensorimotor but not sensorimotor features of human concepts,''
    \emph{Nature Human Behaviour}, 2025.

  \item ``LLMs, Turing Tests and Chinese Rooms,'' \emph{Inquiry}, 2025.

  \item ``The Vector Grounding Problem,'' arXiv:2304.01481, 2023
    (updated 2025).

  \item Quanta Magazine, ``The Converging World of AI,'' January 2026.

  \item Plato, \emph{Republic}, Book VII (c.\ 375 BCE). The allegory of
    the cave.

  \item S.~Kornblith, M.~Norouzi, H.~Lee, and G.~Hinton, ``Similarity
    of Neural Network Representations Revisited,'' \emph{Proc.\ ICML},
    2019.

  \item A.~K.~Zvonkin and L.~A.~Levin, ``The Complexity of Finite
    Objects and the Development of the Concepts of Information and
    Randomness by Means of the Theory of Algorithms,''
    \emph{Russian Mathematical Surveys}, vol.~25, no.~6, pp.~83--124,
    1970.
\end{enumerate}

\end{document}
