\documentclass[11pt,a4paper]{article}
\usepackage{amsmath,amssymb}
\usepackage[margin=2.5cm]{geometry}
\usepackage{enumitem}
\usepackage{hyperref}
\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhf{}
\rhead{ETH Zurich Executive Education}
\lhead{Section 5 --- How Close Are We to AGI?}
\cfoot{\thepage}

\title{Calibration Quiz: How Close Are We to AGI?}
\author{Companion to the ETH Zurich Executive Lecture}
\date{}

\begin{document}
\maketitle

\noindent\textbf{Instructions.}
Answer each question with a single number or multiple choice letter.
For forecasting questions, also write a 90\% confidence interval (a lower and upper
bound you are 90\% sure contains the true answer). This is a calibration
exercise --- the aim is not to be right, but to learn how wide your uncertainty
really is.

\vspace{1em}

\section*{Part I --- Facts}

\begin{enumerate}[label=Q\arabic*.]

\item In Morris et al.\ (DeepMind, 2024), the ``Levels of AGI'' framework
      defines five levels of \emph{performance} (Emerging, Competent, Expert,
      Virtuoso, Superhuman). At which level do the authors place today's
      top frontier models \emph{in the General category}?
\begin{enumerate}[label=(\alph*)]
    \item Emerging AGI \qquad (b) Competent AGI \qquad (c) Expert AGI
    \item Virtuoso AGI \qquad (e) Superhuman AGI (ASI)
\end{enumerate}

\item In Bubeck et al.\ (Microsoft Research, 2023), ``Sparks of Artificial
      General Intelligence,'' which \emph{single} claim is the most
      contested by other researchers?
\begin{enumerate}[label=(\alph*)]
    \item GPT-4 can solve novel mathematical problems
    \item GPT-4 shows signs of a general theory of mind
    \item GPT-4 exhibits behaviour that could reasonably be called
          early AGI
    \item GPT-4 can use tools to interact with the world
\end{enumerate}

\item AlphaGeometry (Trinh et al., \emph{Nature}, 2024) solved how many of
      the 30 most recent IMO geometry problems in the test set?
\begin{enumerate}[label=(\alph*)]
    \item 10 \qquad (b) 18 \qquad (c) 25 \qquad (d) 30
\end{enumerate}

\item AlphaGeometry's training data consists of:
\begin{enumerate}[label=(\alph*)]
    \item 1 million human-annotated geometry proofs
    \item 100 million \emph{synthetic} theorems generated algorithmically
    \item 10 000 IMO-style problems scraped from past competitions
    \item A corpus of geometry textbooks
\end{enumerate}

\item In the METR task-horizon analysis (2024--2026), the time-length of
      tasks that frontier models can complete autonomously at 50\% success
      rate has been doubling approximately every:
\begin{enumerate}[label=(\alph*)]
    \item 2 months \qquad (b) 7 months \qquad
    (c) 18 months \qquad (d) 3 years
\end{enumerate}

\item The Grace et al.\ (2024) survey of AI researchers found that the
      aggregate forecast for a 50\% probability of ``High-Level Machine
      Intelligence'' (HLMI) shifted between the 2022 and 2023 surveys by
      approximately how many years?
\begin{enumerate}[label=(\alph*)]
    \item Later by 13 years \qquad (b) Earlier by 1 year
    \item Earlier by 13 years \qquad (d) No significant change
\end{enumerate}

\item The 2024 \emph{Science} consensus paper ``Managing extreme AI risks
      amid rapid progress'' was co-authored by how many winners of the
      Turing Award?
\begin{enumerate}[label=(\alph*)]
    \item 0 \qquad (b) 1 \qquad (c) 2 \qquad (d) 4
\end{enumerate}

\end{enumerate}

\vspace{1em}
\section*{Part II --- Calibration (give a point estimate AND a 90\% CI)}

\begin{enumerate}[label=Q\arabic*., start=8]

\item In what year will an autonomous AI agent be able to complete,
      on its own, \emph{40 hours} of continuous expert knowledge work at
      a 50\% success rate? (According to a straight-line extrapolation
      of the METR trend.) \\
      Point estimate: \rule{2cm}{0.4pt} \quad
      90\% CI: [\rule{2cm}{0.4pt}, \rule{2cm}{0.4pt}]

\item What fraction of AI researchers in Grace et al.\ (2024) put at least
      a 10\% chance on ``human extinction or similarly permanent and severe
      disempowerment of the human species'' as a consequence of advanced AI? \\
      Point estimate: \rule{2cm}{0.4pt}\%

\item By what year do Grace et al.\ aggregate forecasts assign 50\%
      probability to \emph{full automation of labour}? \\
      Point estimate: \rule{2cm}{0.4pt} \quad
      90\% CI: [\rule{2cm}{0.4pt}, \rule{2cm}{0.4pt}]

\end{enumerate}

\vspace{1em}
\section*{Part III --- Judgement calls (short answers)}

\begin{enumerate}[label=Q\arabic*., start=11]

\item In one sentence: why does Schaeffer et al.\ (NeurIPS 2023) argue that
      many ``emergent abilities'' of LLMs are a \emph{mirage}?

\item In one sentence: what is the single strongest argument Bubeck et al.\
      offer for calling GPT-4 a step towards AGI?

\item In one sentence: what is the single strongest counter-argument a
      skeptic (e.g., Chollet, LeCun, Marcus) would offer?

\item Your company has a 3\,000-person customer-service centre. Based on
      the METR trend, what should your 18-month plan address? Give three
      bullet points.

\item The Bengio--Hinton--Russell consensus paper lists concrete
      \emph{governance} actions. Name at least two.

\end{enumerate}

\vspace{2em}
\hrule
\vspace{1em}

\section*{Answer Key (for the lecturer)}

\begin{enumerate}[label=Q\arabic*.]
\item (a) Emerging AGI --- Morris et al.\ argue current frontier LLMs meet
      the \emph{Emerging} level of General AGI but not Competent.
\item (c) --- the ``sparks of AGI'' framing is what most practitioners pushed
      back on, not the individual capability claims.
\item (c) 25 of 30, vs.\ gold-medallist average of 25.9.
\item (b) 100 million synthetic theorems, generated by a combinatorial
      construction engine + symbolic deduction.
\item (b) $\approx$\,7 months (with a 95\% CI roughly 4--14 months).
\item (c) Earlier by 13 years --- HLMI forecast shifted from 2060 to 2047
      between the 2022 and 2023 rounds.
\item (c) Two: Geoffrey Hinton and Andrew Yao.
\item Under straight-line METR extrapolation from 2024 baseline, 40\,h
      corresponds roughly to the 2030--2032 window; heavy uncertainty is
      warranted.
\item 38\% of respondents gave $\geq 10\%$ probability to extremely bad
      outcomes.
\item The aggregate 50\% forecast for full automation of labour is
      approximately the year 2116, with enormous spread.
\item Schaeffer et al.: many ``emergent'' jumps disappear when you use
      continuous rather than thresholded metrics --- the emergence is an
      artefact of the measurement, not the model.
\item Bubeck et al.: GPT-4's performance is general across an unusually
      wide range of cognitive tasks, not just the ones it was trained on.
\item Skeptics: GPT-4 still fails on simple out-of-distribution tasks that
      a child handles; benchmarks leak into training data; no causal model
      of the world.
\item Acceptable answers include: automate Tier-1 triage, retrain staff
      toward exception handling, instrument the system so failures are
      caught early, prepare the workforce transition.
\item The paper's governance asks include: mandatory capability evaluations
      before deployment, national AI safety institutes, liability for AI
      harms, compute-based licensing for frontier training runs, information
      sharing between developers and regulators.
\end{enumerate}

\end{document}
